---
title: "BellaBeat Report"
author: "Eugene Arkhipov"
date: "2022-10-26"
output: html_document
---

## BellaBeat Data Set

This Data set required to understand the behavior of consumers and their smart devices (Non-BellaBeat). Further, we are required to select one Bellabeat product and apply insights gained.


```{r Setting up Environment, include=FALSE}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("janitor", repos = "http://cran.us.r-project.org")
install.packages("lubridate", repos = "http://cran.us.r-project.org")
install.packages("patchwork", repos = "http://cran.us.r-project.org")
install.packages("rmarkdown", repos = "http://cran.us.r-project.org")


library(tidyverse)
library(janitor)
library(lubridate)
library(ggplot2)
library(patchwork)
library(scales)
library(readr)
library(dplyr)


setwd("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16")

rm(list=ls())
dir("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16", full.names = T)

```

```{r Upload Data, include=FALSE}
activity_daily <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")  
calories_daily <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/dailyCalories_merged.csv")
intensities_daily <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/dailyIntensities_merged.csv")
steps_daily <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/dailySteps_merged.csv")  
heartrate_seconds <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/heartrate_seconds_merged.csv")  
calories_hourly <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv") 
intensities_hourly <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/hourlyIntensities_merged.csv") 
steps_hourly <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/hourlySteps_merged.csv")  
calories_narrow_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteCaloriesNarrow_merged.csv")     
calories_wide_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteCaloriesWide_merged.csv") 
intensities_narrow_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteIntensitiesNarrow_merged.csv")
intensities_wide_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteIntensitiesWide_merged.csv")
mets_narrow_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteMETsNarrow_merged.csv") 
sleep_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteSleep_merged.csv") 
steps_narrow_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteStepsNarrow_merged.csv") 
steps_wide_minute <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/minuteStepsWide_merged.csv") 
sleep_day <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
weight_loginfo <- read.csv("C:/Users/mader/OneDrive/Desktop/Capstone Data/Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv")
```


```{r Check how many distinct ids the data set has, include=FALSE}
n_distinct(activity_daily$Id)
n_distinct(calories_daily$Id)
n_distinct(calories_hourly$Id)
n_distinct(calories_narrow_minute$Id)
n_distinct(calories_wide_minute$Id)
n_distinct(heartrate_seconds$Id)
n_distinct(intensities_daily$Id)
n_distinct(intensities_hourly$Id)
n_distinct(intensities_narrow_minute$Id)
n_distinct(intensities_wide_minute$Id)
n_distinct(mets_narrow_minute$Id)
n_distinct(sleep_day$Id)
n_distinct(sleep_minute$Id)
n_distinct(steps_daily$Id)
n_distinct(steps_hourly$Id)
n_distinct(steps_narrow_minute$Id)
n_distinct(steps_wide_minute$Id)
n_distinct(weight_loginfo$Id)
```

```{r summarize data, include=FALSE}
summary(activity_daily)
summary(calories_daily)
summary(calories_hourly)
summary(calories_narrow_minute)
summary(heartrate_seconds)
summary(intensities_daily)
summary(intensities_hourly)
summary(intensities_narrow_minute)
summary(mets_narrow_minute)
summary(sleep_day)
summary(sleep_minute)
summary(steps_daily)
summary(steps_hourly)
summary(steps_narrow_minute)
summary(weight_loginfo)
```


```{r check for null values in ever column, include=FALSE}
is.null(activity_daily)
is.null(calories_daily)
is.null(calories_hourly)
is.null(calories_narrow_minute)
is.null(heartrate_seconds)
is.null(intensities_daily)
is.null(intensities_hourly)
is.null(intensities_narrow_minute)
is.null(mets_narrow_minute)
is.null(sleep_day)
is.null(sleep_minute)
is.null(steps_daily)
is.null(steps_hourly)
is.null(steps_narrow_minute)
is.null(weight_loginfo)
```

```{r # add data columns, total active minutes, include=FALSE}
activity_daily <- activity_daily %>% 
  mutate(total_active_minutes = VeryActiveMinutes+FairlyActiveMinutes+LightlyActiveMinutes)

intensities_daily <- intensities_daily %>% 
  mutate(total_active_minutes = VeryActiveMinutes+FairlyActiveMinutes+LightlyActiveMinutes)
```

```{r total minutes worn, hours worn,  include=FALSE}
## total minutes worn
activity_daily <- activity_daily %>% 
  mutate(total_minutes_worn = VeryActiveMinutes+FairlyActiveMinutes+LightlyActiveMinutes+SedentaryMinutes)

## total hours worn
activity_daily$total_hours_worn <- activity_daily$total_active_minutes / 60
```
 
```{r format Data Types (id to character, minutes to mins), make time columns identical format, include=FALSE}
activity_daily_formatted <- mutate(activity_daily, Id = as.character(Id),
                                   ActivityDate=as_date(ActivityDate, format=("%m/%d/%Y")))
                               

calories_daily_formatted <-  mutate(calories_daily, Id = as.character(Id),
                                    ActivityDate=as_date(ActivityDay, format=("%m/%d/%Y")))


intensities_daily_formatted <- mutate(intensities_daily, Id = as.character(Id),
                                      ActivityDate=as_date(ActivityDay, format=("%m/%d/%Y")))
                                   


sleep_day_formatted <-  mutate(sleep_day, Id = as.character(Id),
                               ActivityDate=as_date(SleepDay, format=("%m/%d/%Y")))
                     

weight_loginfo_formatted <- mutate(weight_loginfo, Id=as.character(Id),
                                   ActivityDate=as_date(Date, format=("%m/%d/%Y")))

steps_daily_formatted <- mutate(steps_daily, Id=as.character(Id),
                                ActivityDate=as_date(ActivityDay, format=("%m/%d/%Y")))
```

```{r Rename time/date column for consistency, include=FALSE}
sleep_day <- rename (sleep_day,
                     ActivityDate=SleepDay)

calories_daily <- rename(calories_daily, 
                         ActivityDate=ActivityDay)

intensities_daily <- rename(intensities_daily, 
                            ActivityDate=ActivityDay)

weight_loginfo <- rename(weight_loginfo, 
                         ActivityDate=Date)
```


```{r join into one frame, include=FALSE}
activity_sleep_combined <- full_join(activity_daily_formatted,sleep_day_formatted, by=c("Id","ActivityDate"))

activity_full <- full_join(activity_sleep_combined, weight_loginfo_formatted, by=c("Id","ActivityDate"))

```

```{r sumarize new data set, include=FALSE}
View(activity_full)
str(activity_full)
summary(activity_full)
```

```{r create summary rows/table, customer id, total minutes active, total minutes asleep, amount of days worn, sedentary minutes, total steps per day, echo=TRUE, include=FALSE}
activity_count <- activity_daily_formatted %>% 
  group_by(Id) %>% 
  summarize(count=n())

average_steps <- steps_daily_formatted %>% 
  group_by(Id) %>% 
  summarize(mean(StepTotal))

average_active_minutes <- intensities_daily_formatted %>% 
  group_by(Id) %>% 
  summarize(mean(total_active_minutes))

average_sedentary_minutes <- intensities_daily_formatted %>% 
  group_by(Id) %>% 
  summarize(mean(SedentaryMinutes))


average_calories <- calories_daily_formatted %>% 
  group_by(Id) %>% 
  summarize(mean(Calories))

average_weight <- weight_loginfo_formatted %>% 
  group_by(Id) %>% 
  summarize(mean(WeightPounds))

average_bmi <- weight_loginfo_formatted %>% 
  group_by(Id) %>% 
  summarize(mean(BMI))
```

```{r Create dataframe with new summary rows, include=FALSE}
summary_table <- list(activity_count,
                      average_steps,
                      average_active_minutes,
                      average_sedentary_minutes,
                      average_calories,
                      average_weight,
                      average_bmi) %>% 
  reduce(left_join, by='Id')                            
```

```{r rename summary_table columns, include=FALSE}

summary_table <- rename (summary_table,
                     average_steps="mean(StepTotal)")

summary_table <- rename (summary_table,
                         average_active_minutes="mean(total_active_minutes)")

summary_table <- rename (summary_table,
                         average_sedentary_minutes="mean(SedentaryMinutes)")

summary_table <- rename (summary_table,
                         average_calories="mean(Calories)")

summary_table <- rename (summary_table,
                         average_weight="mean(WeightPounds)")

summary_table <- rename (summary_table,
                         average_bmi="mean(BMI)")
```

```{r average minutes column, include=FALSE}
summary_table$total_average_minutes  <- rowSums(summary_table[ , c(4,5)])
```

## Summary of Averages

- Avg steps = 7519
- Avg Active Hours = 3.75
- Avg sedentary Hours = 16.7
- Avg Calories = 2282

```{r summarise summary_table, message=FALSE, warning=FALSE, include=FALSE}
summary_table %>% 
  summarize(avg_steps=mean(average_steps), avg_active_minutes=mean(average_active_minutes),
avg_sedentary_minutes=mean(average_sedentary_minutes), avge_calories = mean(average_calories), 
avg_weight=mean(average_weight), avg_bmi = mean(average_bmi), 
total_avg_hours=mean(total_average_minutes))
```            


```{r convert to hours, message=FALSE, warning=FALSE, include=FALSE}
summary_avg_hours <- summary_table %>% 
  summarize(avg_steps=mean(average_steps), avg_active_hrs=mean(average_active_minutes / 60),
            avg_sedentary_hrs=mean(average_sedentary_minutes / 60), avg_calories = mean(average_calories), 
            avg_weight=mean(average_weight), avg_bmi = mean(average_bmi), 
            total_avg_hrs=mean(total_average_minutes / 60)) %>% 
  print()
```


## Preliminary Observations

- Weight/BMI data was Incomplete/Missing 
- 33 Unique consumers
- Only 8 provided BMI/Weight Data
- Only 14 provided heartrate Data
- Only 24 provided sleep Data
- "Active time" is a sum of Very Active, Fairly Active, and Lightly Active people

```{r visualization, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=summary_table, aes(x=average_steps, y=Id))+geom_bar(stat="identity")
ggplot(data=summary_table, aes(x=average_active_minutes, y=Id))+geom_bar(stat="identity")
ggplot(data=summary_table, aes(x=average_sedentary_minutes, y=Id))+geom_bar(stat="identity")
ggplot(data=summary_table, aes(x=average_calories, y=Id))+geom_bar(stat="identity")
ggplot(data=summary_table, aes(x=average_weight, y=Id))+geom_bar(stat="identity")
ggplot(data=summary_table, aes(x=average_bmi, y=Id))+geom_bar(stat="identity")
```

```{r pivot longer to plot bar chart, echo=FALSE, message=FALSE, warning=FALSE}
summary_pivot <- summary_table %>% 
  pivot_longer(cols=c("average_sedentary_minutes", "average_active_minutes"),
               names_to="time_type",
               values_to="minutes")
```

```{r time comparison for all users, echo=FALSE}
ggplot(summary_pivot, aes(x=Id, y=minutes, fill=time_type))+
  geom_col(position = "dodge")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

ggplot(summary_pivot, aes(x=minutes, fill=time_type, y=Id))+
  geom_col()+
  geom_text(aes(label=round(minutes),size=3))
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Deductions

- There is a large difference between active time and sedentary time, more than twice as much sedentary time as all types of active time combined
- Manually input of data produces incomplete results (weight) because not everyone inputs BMI/Weight Data
- 9 people take their devices off during sleep (for comfort or to recharge)
- Most People are under 8,000 steps per day
- Average Hours of sleep/night for all consumers is about 7 Hours


## Recommendations
- Motivate users to input weight data
- Motivate users to get in the habit of going to sleep for time
- Present users with a pair of devices so that one can charge while the other is being worn to minimize data omission - Motivate users to be more active/not sedentary (any time of activity)
- Motivation can be through notifications, vibrations, reminders, goals/targets on the connected app